Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model

### 2025-10-02 paper/paper.qmd
SD model

---
title: Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: ICR
    longname: interquartile range
---

@@

```{r setup, include=FALSE}
# here::i_am("paper/index.qmd")
run_sims <- FALSE

@@

The transmission of pathogenic bacteria in hospitals results in excessive morbidity and mortality.
Relevant examples include *Clostridioides difficile*, \acr{VRE}, \acr{ESBL}-producing Enterobacerales, and \acr{CPE}.
Drug resistance further compounds the problem --- for example, \acr{CPE} bloodstream infection is associated with an increased risk of in-hospital mortality (risk difference of 0.25) compared to carbapenem-susceptible infection [@budhramHealthOutcomesAttributable2020].
\acr{CDI} is associated with prolonged hospital stay and increased healthcare costs, causing a significant operational burden [@forsterEffectHospitalacquiredInfection2012].
Furthermore, the therapeutic antimicrobial options available for \acr{VRE}, \acr{ESBL}, and \acr{CPE} are limited, leading to increased use of \acr{WHO} reserve agents, which further exacerbates the problem of antimicrobial resistance [@organization2019WHOAWaRe2019].

\acr[case=sentence]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings. Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected/colonised patients [@worldhealthorganizationGlobalReportInfection2022].
A key role of \acr{IPC} teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected/colonised patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
The impact of these interventions is often difficult to quantify.
Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable.
Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such rates of antimicrobial usage, and to structural differences such as hospital layout.

@@

Here, we describe an agent-based model that describes the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods. The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}.

@@

$$
x_p(t) \epsilon \{0, 1\}
$$

@@

-   $x_p(t) = 0$ if patient $p$ is not infected at time $t$,
-   $x_p(t) = 1$ if patient $p$ is infected at time $t$.

@@

$$ 
P_\text{de novo}(t) = 1 - \text{Exp} {( - r_0 T(t) )},
$$

@@

$$
P_\text{toilet}(t) = 1 - \text{Exp} ( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} )
$$

@@

-   $T(t)$ is a binary indicator for antimicrobial therapy:

@@

-   $U_i(t)$ is the number of times patient $i$ frequents the toilet within the time interval $t$,

-   $C(t_j)$ is the contamination level of the toilet at time $t_j$.

@@

The results of the trials were combined to determine the outcome for patient $i$ at time $t$:

$$
X_i(t) = \text{Max}(X_\text{de novo}, X_\text{toilet}),
$$

where $X_i(t)$ = 1 if patient $i$ becomes infected at time $t$, and 0 otherwise.

@@

$$
\frac{dS}{dt} = pA - D \cdot \frac{S}{S + I} - \beta S - rS
$$

$$
\frac{dI}{dt} = (1 - p)A - D \cdot \frac{I}{S + I} + \beta S + rS
$$

@@

-   $dS/dt$ is the rate of change of susceptible/non-infected inpatients,
-   $S$ is the number of susceptible inpatients,
-   $I$ is the number of infected/colonised inpatients,
-   $A$ is the rate of new admissions,
-   $p$ is the proportion of new admissions that are susceptible/non-infected, while $(1 - p)$ is community colonisation/infection rate,
-   $\beta$ is the transmission coefficient,
-   $r$ is the rate of random (*de novo*) colonisation/infection,
-   $D$ is the rate of patient discharges.

@@

@@

| Name | Distribution | Parameters | Description | Reference |
|---------------|---------------|---------------|---------------|---------------|
| Baseline (pre-outbreak) antimicrobial prescription rate | Constant | 49.9% | Proportion of patients administered antimicrobials during admission | @versportenAntimicrobialConsumptionResistance2018 |
| Admission duration | Poisson | $\lambda = 4.6  \text{ days}$ |  | @levitkryankelixhauserastrangesekassedccoffeyrHCUPFactsFigures2007 |
| Proportion of toilets that are shared between \>1 patients | Constant | 60% | The remaining proportion (40%) of toilets were dedicated to one patient | N/A |
| Community colonisation rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
| Antimicrobial effect | Uniform | $ a = 1.31, b = 1.87 $ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
| CDI incidence | Uniform | $a=2.8, b=12.1$ | Per 10,000 patient-days | @marraIncidenceOutcomesAssociated2020 |
| Mean toilet frequenting rate | Uniform | $a=0.5,b=5$ | Per day; used as \lambda parameter for Poisson distribution |  |
| Mean toilet cleaning rate | Uniform | $a=0.5, b=3$ | Per day; used as \lambda parameter for Poisson distribution |  |
| Toilet contamination effect | Uniform | $a=0.01,b=0.99$ | Contamination impact by infected/colonised patients through toilet use |  |
| Toilet cleaning effect | Uniform | $a=0.01, b=0.99$ | Impact of toilet cleaning on contamination |  |

@@

```{r fig-posterior-densities, include=TRUE}
#| eval: true
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| warning: false
#| fig-cap: "Changes in parameters during outbreak and enhanced infection control periods. Density plots show the posterior distributions of key simulation parameters after calibration using approximate Bayesian computation."
#| fig-subcap:
#|   - "General parameters"
#|   - "Toilet-related parameters"
#| layout-ncol: 1
abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(!contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~name, scales = "free_x") +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "green", "Outbreak" = "red", "Control" = "blue"))

abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~name, scales = "free_x") +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "green", "Outbreak" = "red", "Control" = "blue"))
```

Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-posterior-predictive-check). In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_posterior_predictive_sims_rates$rate_median, 2)` cases per 1000 patient days \[\acr{ICR} range `r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\]. During the outbreak period, simulations had a median rate of `r round(outbreak_uniform_sims_rates$rate_median, 2)` \[\acr{IQR} `r round(outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(outbreak_uniform_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[\acr{IQR} 3.32--4.41\] (see @fig-posterior-predictive-check).

@@

@@

>>>>

---
title: Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: IQR
    longname: interquartile range
  insert_loa: false
---

@@

```{r setup, include=FALSE}
# here::i_am("paper/index.qmd")
run_sims <- FALSE
set.seed(42)

@@

The transmission of pathogenic bacteria in hospitals results in excessive morbidity and mortality.
Relevant examples include *Clostridioides difficile*, \acr{VRE}, \acr{ESBL}-producing Enterobacerales, and \acr{CPE}.
Drug resistance further compounds the problem --- for example, \acr{CPE} bloodstream infection is associated with a considerably increased risk of in-hospital mortality compared to carbapenem-susceptible infection [@budhramHealthOutcomesAttributable2020].
\acr{CDI} is associated with prolonged hospital stay and increased healthcare costs, causing a significant operational burden [@forsterEffectHospitalacquiredInfection2012].
Furthermore, the therapeutic antimicrobial options available for \acr{VRE}, \acr{ESBL}, and \acr{CPE} are limited, leading to increased use of \acr{WHO} reserve agents, which further exacerbates the problem of antimicrobial resistance [@organization2019WHOAWaRe2019].

\acr[case=sentence]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings. Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected/colonised patients [@worldhealthorganizationGlobalReportInfection2022].
A key role of \acr{IPC} teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected/colonised patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
The impact of these interventions is often difficult to quantify.
Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable.
Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such as rates of antimicrobial usage, and to structural differences such as hospital layout.

@@

Here, we describe an agent-based model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods. The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}.

@@

$$
x_p(t) \in \{0, 1\}
$$

@@

-   $t \in \mathbb{N}_0 = \{0, 1, 2, \ldots\}$ is the time step (in days),
-   $x_p(t) = 0$ if patient $p$ is not infected at time $t$,
-   $x_p(t) = 1$ if patient $p$ is infected at time $t$.

@@

$$ 
P_\text{de novo}(t) = 1 - \text{exp}{ {( - r_0 T(t) )} },
$$

@@

$$
P_\text{toilet}(t) = 1 - \text{exp}{ \left( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} \right) }
$$

@@

-   $U_i(t)$ is the number of times patient $p$ frequents the toilet within the time interval $t$,

-   $C(t_j)$ is the contamination level of the toilet at time $t_j$, and

@@

@@

The results of the trials were combined to determine the outcome for patient $p$ at time $t$; the patient became infected if either trial was sucessful:

$$
X_p(t) = \text{Max}(X_\text{de novo}, X_\text{toilet}),
$$

where $X_i(t)$ = 1 if patient $p$ becomes infected at time $t$, and 0 otherwise.

## Deterministic model

@@

$$
\frac{dS}{dt} = (1-p)A - D \cdot \frac{S}{S + I} - \beta S - rS
$$

$$
\frac{dI}{dt} = pA - D \cdot \frac{I}{S + I} + \beta S + rS
$$

@@

-   $dS/dt$ is the rate of change of susceptible/non-infected inpatients,
-   $S$ is the number of susceptible inpatients,
-   $I$ is the number of infected/colonised inpatients,
-   $A$ is the rate of new admissions,
-   $p$ is the community colonisatio rate,
-   $\beta$ is the transmission coefficient,
-   $r$ is the rate of random (*de novo*) colonisation/infection,
-   $D$ is the rate of patient discharges.

@@

![System dynamics model](images/sd_model.svg)

@@

| Name | Distribution | Parameters | Description | Reference |
|---------------|---------------|---------------|---------------|---------------|
| Baseline (pre-outbreak) antimicrobial prescription rate | Constant | 49.9% | Proportion of patients administered antimicrobials during admission | @versportenAntimicrobialConsumptionResistance2018 |
| Admission duration | Poisson | $\lambda = 4.6  \text{ days}$ |  | @levitkryankelixhauserastrangesekassedccoffeyrHCUPFactsFigures2007 |
| Proportion of toilets that are shared between \>1 patients | Constant | 60% | The remaining proportion (40%) of toilets were dedicated to one patient | N/A |
| Community colonisation rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
| Antimicrobial effect | Uniform | $a = 1.31, b = 1.87$ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
| CDI incidence | Uniform | $a=2.8, b=12.1$ | Per 10,000 patient-days | @marraIncidenceOutcomesAssociated2020 |
| Mean toilet frequenting rate | Uniform | $a=0.5,b=5$ | Per day; used as \lambda parameter for Poisson distribution |  |
| Mean toilet cleaning rate | Uniform | $a=0.5, b=3$ | Per day; used as \lambda parameter for Poisson distribution |  |
| Toilet contamination effect | Uniform | $a=0.01,b=0.99$ | Contamination impact by infected/colonised patients through toilet use |  |
| Toilet cleaning effect | Uniform | $a=0.01, b=0.99$ | Impact of toilet cleaning on contamination |  |

@@

```{r fig-posterior-densities, include=TRUE}
#| eval: true
#| echo: false
#| fig.width: 18
#| fig.height: 18
#| warning: false
#| fig-cap: "Changes in parameters during outbreak and enhanced infection control periods. Density plots show the posterior distributions of key simulation parameters after calibration using approximate Bayesian computation."
general_params <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(!contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type, linetype = type)) +
  geom_density(alpha = .3) +
  facet_wrap(~name, scales = "free_x") +
  theme_minimal() +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "#66c2a5", "Outbreak" = "#fc8d62", "Control" = "#8da0cb")) +
  scale_linetype_manual(values = c("Baseline" = "solid", "Outbreak" = "dashed", "Control" = "dotted")) +
  # remove gridlines
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

toilet_params <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type, linetype = type)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~name, scales = "free_x") +
  theme_minimal() + 
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "#66c2a5", "Outbreak" = "#fc8d62", "Control" = "#8da0cb")) +
  scale_linetype_manual(values = c("Baseline" = "solid", "Outbreak" = "dashed", "Control" = "dotted")) +
  # remove gridlines
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

# combine plots
library(patchwork)
general_params / toilet_params + plot_layout(heights = c(0.5, 1))
```

Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-posterior-predictive-check). In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_posterior_predictive_sims_rates$rate_median, 2)` cases per 1000 patient days \[\acr{IQR} range `r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[\acr{IQR} 0.94--1.56\]. During the outbreak period, simulations had a median rate of `r round(outbreak_uniform_sims_rates$rate_median, 2)` \[\acr{IQR} `r round(outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(outbreak_uniform_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[\acr{IQR} 3.32--4.41\] (see @fig-posterior-predictive-check).

@@

```{r}
stochastic_sd <- read_csv("../out/sd_stochastic experiment-table.csv", skip = 6)
```

```{r}
Rcpp::sourceCpp("../models/sd_stochastic.cpp")
```

```{r}
set.seed(42)
randomLHS_sd <- randomLHS(1000000, 8)
sd_param_grid <- data.frame(init_population = rep(610, 1000),
                            A = 68,
                            r = qunif(randomLHS_sd[, 3], 0.001, 0.01),
                            baseline_beta = qunif(randomLHS_sd[, 4], 0.0001, 0.01),
                            baseline_p = qunif(randomLHS_sd[, 5], 0.001, 0.1),
                            outbreak_beta = qunif(randomLHS_sd[, 6], 0.0001, 0.01),
                            outbreak_p = qunif(randomLHS_sd[, 7], 0.001, 0.1),
                            outbreak_start = rep(1042, 1000),
                            outbreak_end = rep(1234, 1000),
                            n_days = rep(2585, 1000))

sd_param_grid <- sd_param_grid %>% 
  filter(baseline_beta < outbreak_beta & baseline_p < outbreak_p) %>% 
  slice_sample(n = 10000) %>% 
  mutate(param_row = row_number())
n_seeds <- 2
sd_sims <- 
  pmap_dfr(sd_param_grid,
        \(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days, param_row) {
          map(1:n_seeds,
              \(seed) {
                set.seed(seed)
                sd_stochastic(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days) %>% 
                  mutate(`random-seed` = seed,
                         param_row = param_row)
              }) %>% 
            bind_rows()
        }, .progress = T)

sd_sims <- sd_sims %>% 
  filter(day > 31) %>% 
  group_by(param_row, `random-seed`) %>% 
  # add dates
  mutate(date_sim = seq(from = ymd("2002-01-01"),
                        by = "1 day",
                        length.out = n()))

# rates by month
sd_sims_rates <- sd_sims %>% 
  group_by(param_row, `random-seed`) %>% 
  mutate(year = year(date_sim),
         month = month(date_sim)) %>%
  group_by(param_row, `random-seed`, year, month) %>%
  # calculate rate of hosp_cases per 1000 patient days
  summarise(rate = (sum(hosp_cases) / sum(S + I)) * 1000,
            date_sim = first(date_sim))

# collapse to mean rates by seed
sd_sims_rates_mean <- sd_sims_rates %>% 
  group_by(param_row, year, month) %>%
  summarise(rate = mean(rate),
            date_sim = first(date_sim))

# pivot wider to get ready for abc

sd_sumstats <- sd_sims_rates_mean %>% 
  ungroup() %>% 
  dplyr::select(-year, -month) %>%
  filter(date_sim <= max(salgado$x)) %>% 
  pivot_wider(names_from = date_sim, values_from = rate)

# abc
abc_sd <- abc(salgado$rates,
              sd_param_grid, dplyr::select(sd_sumstats, -param_row), tol=0.0025, method='rejection')

# posterior predictive checks
posterior_check_params <- abc_sd$unadj.values %>% 
  as_tibble() %>% 
  slice_sample(n = 100, replace = TRUE) %>% 
  mutate(seed = sample(1:1000, n(), replace = FALSE))

posterior_predictive_sim_sd <- 
  pmap_dfr(posterior_check_params,
        \(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days, param_row, seed) {
            set.seed(seed)
            sd_stochastic(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days) %>% 
              mutate(`random-seed` = seed,
                     param_row = param_row) %>% 
              bind_rows()
        }, .progress = T)

posterior_predictive_sim_sd <- posterior_predictive_sim_sd %>%
  filter(day > 31) %>% 
  group_by(`random-seed`) %>% 
  # add dates
  mutate(date_sim = seq(from = ymd("2002-01-01"),
                        by = "1 day",
                        length.out = n()))

posterior_predictive_sim_sd_rates <- posterior_predictive_sim_sd %>%
  group_by(`random-seed`) %>% 
  mutate(year = year(date_sim),
         month = month(date_sim)) %>%
  group_by(`random-seed`, year, month) %>%
  # calculate rate of hosp_cases per 1000 patient days
  summarise(rate = (sum(hosp_cases) / sum(S + I)) * 1000,
            date_sim = first(date_sim))

@@

```
```{r fig-abc-sd-posteriors, include=TRUE}
#| echo: false
#| fig.cap: "Posterior distributions of parameters obtained from approximate Bayesian computation using summary statistics generated from the stochastic deterministic model. The red line shows the prior distribution; the blue line shows the posterior distribution."
```

```{r fig-abc-sd-simulated-rates, include=TRUE}
#| echo: false
#| fig.cap: "Simulated rates of _C. difficile_ infection generated from the simulations retained after applying the rejection approximate Bayesian computation algorithm to the stochastic deterministic model. The red line shows the true observed infection rates. The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."

abc_sd$ss %>% as.matrix %>% 
  as.data.frame() %>% 
  pivot_longer(cols = everything(), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(date_sim) %>%
  summarise(m = median(rate),
            q_high = quantile(rate, 0.75),
            q_low = quantile(rate, 0.25),
            .groups = "keep") %>%
  ggplot() +
  geom_line(aes(x = date_sim, y = m)) +
  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
  guides(color="none") +
  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
```

```{r fig-abc-sd-posterior-predictive-check, include=TRUE}
#| echo: false
#| fig.cap: "Posterior predictive check of the stochastic deterministic model. Simulations were generated by sampling with replacement from the posterior parameter distributions after approximate Bayesian computation. Each simulation was run using different random number generators. The red line shows the true observed infection rates. The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."

posterior_predictive_sim_sd_rates %>% 
  pivot_wider(names_from = date_sim, values_from = rate) %>% 
  pivot_longer(cols = !c(`random-seed`, `param_row`), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(date_sim) %>%
  summarise(m = median(rate),
            q_high = quantile(rate, 0.75),
            q_low = quantile(rate, 0.25),
            .groups = "keep") %>%
  ggplot() +
  geom_line(aes(x = date_sim, y = m)) +
  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
  guides(color="none") +
  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
